Precise spiking motifs in neurobiological and neuromorphic data

### 2022-11-15 manuscript.tex
Update on Overleaf.

% GAP
However, computational neuroscience models have suggested that the precise timing within a sequence of spikes may play a crucial role, and that neurons may be as well integrators than synchrony detectors~\citep{abeles_role_1982}. We will investigate this very hypothesis in this review. Indeed, the response of one biological neuron largely depends on the precise timing of the sequence of incoming spikes that reach its soma~\citep{paugam-moisy_computing_2012}. In comparison to a classical analogical vector of inputs, this \emph{event-based representation} observed in the neural code is essential in understanding information processing~\citep{carr_processing_1993}. For instance, it expands the capabilities of analog representations (which rely on the firing rate) by considering a representation based on repetitions of spiking motifs at precise times of occurrence. This hypothesis is directly inspired by neurobiological observations and expands the capabilities of analog representations (which are based on spiking firing rates) by considering a representation based on instances of spiking motifs at precise times of occurrence. Additionally, numerous studies demonstrated the importance of precise timing in neural population activity~\citep{davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{perrinet_coding_2004,gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2003,carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and further suggest the existence of precise spatio-temporal spiking motifs in the input which excites neurons.  A mathematical formalization would be particularly well suited to neuromorphic computing~\citep{roy_towards_2019}, and would allow for the supervised or self-supervised learning of such motifs in any event-driven data.  Crucially, validating this hypothesis would also be crucial in our understanding of neural processes. %
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Dynamics of vision and consequences on the neural code}%Precise timing in the neural code enables efficient vision}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Let us start with a focus of the state-of-the-art on the role of dynamics in vision. Broadly speaking, vision is the set of processes that allow us to make sense of the luminous world, and is an intensively studied field in neuroscience, in particular with respect to deciphering the neural code. In most mammals, light enters the eye to induce neural activity on the retina, which maintains a certain similitude between the topology of external visual space and its representation on the retina, called retinotopy. The origins of this neuroscientific question can be found in the first experiments from Pierre Flourens which, using lesions in animals, demonstrated the relationship between visual sensations and activity in the cerebral cortex~\citep{flourens_recherches_1842,pearce_marie-jean-pierre_2009}. This was also observed by Lord Adrian when recording the activity the frog's visual system~\citep{adrian_impulses_1926}. In a series of seminal studies,~\citet{hubel_receptive_1968} showed that this activity could be selective to different features, such as the visual orientation or motion of elementary contours. For a large proportion of neurons, there is remarkably a monotonic relationship between the contrast of visual features and the firing frequency of neurons. However, there is no consensus to explain the multiple nonlinear mechanisms that transform the visual scene into retinotopic neural activity maps, even though these processes seem to constitute essential pieces to this puzzle~\citep{carandini_normalization_2012}.

@@

% retina
How may information about the sensory world be encoded in the spiking pattern?  At the level of the retina, it has been shown that a coding of luminance values in the image using the timing of the spikes may be at work~\citep{gollisch_rapid_2008}. In particular, these results show that the response of ganglion cells to the visual gratings projected on the retina could be encoded in the latency of the response and not only in the frequency of the discharge, as it is often assumed. These results have been extended to natural images and show a qualitatively similar behavior. The authors' conclusion is that the precise spiking latency of the neurons encodes the spatial features of the image. Interestingly, such a precise latency mechanism may underlie some visual illusions, e.g., the false color illusion in the \href{https://michaelbach.de/ot/col-Benham/index.html}{Benham Top} based on center-surround interactions in the parvocellular pathway~\citep{kenyon_theory_2004}. This evidence found in the retina can be extended to other areas such as the visual cortex.
%----------------------------%
% C_MainenSejnowski1995_Perrinet.ipynb
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/replicating_MainenSejnowski1995.pdf} 
\caption{The timing of the spikes produced following the repetition of a step stimuli is less reproducible than that to a noisy stimulus. The stimulus is shown in the top row, while trial repetitions of a simulated Leaky Integrate-and-Fire neuron are shown in the second row (with membrane potential represented by dark blue color when light yellow colors when depolarized) and quantified by the average firing rate across trials (lower row). While this seems paradoxical at first sight, this is due to the use of the same \emph{frozen} noise at each repetition and highlight the highly reproducible pattern of spikes when it is driven by a highly dynamic input. See this~\href{https://github.com/laurentperrinet/2022_UE-neurosciences-computationnelles/blob/master/C_MainenSejnowski1995_Perrinet.ipynb}{notebook} for a replication of this result using a simple LIF model.}\label{fig:mainen}
\end{figure}
%----------------------------%

@@

    \caption{\textbf{Core mechanism of polychrony detection~\citep{izhikevich_polychronization_2006}.} \textit{(Left)}~In that example, three presynaptic neurons denoted \textit{b}, \textit{c} and, \textit{d} are fully connected to two post-synaptic \textit{a} and \textit{e}, with different delays of respectively $1$, $5$, and $9~\ms$ for \textit{a} and  $8$, $5$, and $1~\ms$ for \textit{e}. \textit{(Middle)}~if a synchronous pulse is emitted in all presynaptic neurons, this will generate post-synaptic potentials at these different delays, and these may not be sufficient to generate a spike in either neuron. \textit{(Right)}~If a sequence of activation is generated in the presynaptic neurons such that, taking into account the heterogeneous delays, these converge at the same time on neuron \textit{a} (here, at $t=10~\ms$) then a spike is emitted (here in red color), while none is emitted on neuron \textit{e}.

>>>>

% GAP
However, computational neuroscience models have suggested that the precise timing within a sequence of spikes may play a crucial role, and that neurons may be as well integrators than synchrony detectors~\citep{abeles_role_1982}. We will investigate this very hypothesis in this review. Indeed, the response of one biological neuron largely depends on the precise timing of the sequence of incoming spikes that reach its soma~\citep{paugam-moisy_computing_2012}. In comparison to a classical analogical vector of inputs, this \emph{event-based representation} observed in the neural code is essential in understanding information processing~\citep{carr_processing_1993}. For instance, it expands the capabilities of analog representations (which rely on the firing rate) by considering a representation based on repetitions of spiking motifs at precise times of occurrence. This hypothesis is directly inspired by neurobiological observations and expands the capabilities of analog representations (which are based on spiking firing rates) by considering a representation based on instances of spiking motifs at precise times of occurrence. Additionally, numerous studies demonstrated the importance of precise timing in neural population activity~\citep{davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{perrinet_coding_2004,gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2003,carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and further suggest the existence of precise spatio-temporal spiking motifs in the input which excites neurons.  A mathematical formalization would be particularly well suited to neuromorphic computing~\citep{roy_towards_2019}, and would allow for the supervised or self-supervised learning of such motifs in any event-driven data.  Crucially, validating this hypothesis would also be crucial in our understanding of neural processes. %
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Dynamics of vision and consequences on the neural code}%Precise timing in the neural code enables efficient vision}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Let us start with a focus on the state-of-the-art of the role of dynamics in vision. Broadly speaking, vision is the set of processes that allow us to make sense of the luminous world, and is an intensively studied field in neuroscience, in particular with respect to deciphering the neural code. In most mammals, light enters the eye to induce neural activity on the retina, which maintains a certain similitude between the topology of external visual space and its representation on the retina, called retinotopy. The origins of this neuroscientific question can be found in the first experiments from Pierre Flourens which, using lesions in animals, demonstrated the relationship between visual sensations and activity in the cerebral cortex~\citep{flourens_recherches_1842,pearce_marie-jean-pierre_2009}. This was also observed by Lord Adrian when recording the activity the frog's visual system~\citep{adrian_impulses_1926}. In a series of seminal studies,~\citet{hubel_receptive_1968} showed that this activity could be selective to different features, such as the visual orientation or motion of elementary contours. For a large proportion of neurons, there is remarkably a monotonic relationship between the contrast of visual features and the firing frequency of neurons. However, there is no consensus to explain the multiple nonlinear mechanisms that transform the visual scene into retinotopic neural activity maps, even though these processes seem to constitute essential pieces to this puzzle~\citep{carandini_normalization_2012}.

@@

% retina
How may information about the sensory world be encoded in the spiking pattern?  At the level of the retina, it has been shown that a coding of luminance values in the image using the timing of the spikes may be at work~\citep{gollisch_rapid_2008}. In particular, these results show that the response of ganglion cells to the visual gratings projected on the retina could be encoded in the latency of the response and not only in the frequency of the discharge, as it is often assumed. These results have been extended to natural images and show a qualitatively similar behavior. The authors' conclusion is that the precise spiking latency of the neurons encodes the spatial features of the image. Interestingly, such a precise latency mechanism may underlie some visual illusions, e.g., the false color illusion in the \href{https://michaelbach.de/ot/col-Benham/index.html}{Benham Top} based on center-surround interactions in the parvocellular pathway~\citep{kenyon_theory_2004}. This evidence found in the retina can be extended to other areas such as the visual cortex.
%----------------------------%
% C_MainenSejnowski1995_Perrinet.ipynb
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/replicating_MainenSejnowski1995.pdf} 
\caption{The timing of the spikes produced following the repetition of a step stimulus is less reproducible than that to a noisy stimulus. The stimulus current value over time for a step stimulus (top left) and for a noisy one (top right). Trial repetitions of a Leaky Integrate-and-Fire neuron stimulated by the stimulus on the upper row (middle row). Membrane potential represented by dark blue color when light yellow colors when depolarized) and quantified by . The average firing rate across trials (lower row). While this seems paradoxical at first sight, this is due to the use of the same \emph{frozen} noise at each repetition and highlight the highly reproducible pattern of spikes when it is driven by a highly dynamic input. See this~\href{https://github.com/laurentperrinet/2022_UE-neurosciences-computationnelles/blob/master/C_MainenSejnowski1995_Perrinet.ipynb}{notebook} for a replication of this result using a simple LIF model.}\label{fig:mainen}
\end{figure}
%----------------------------%

@@

    \caption{\textbf{Core mechanism of polychrony detection~\citep{izhikevich_polychronization_2006}.} \textit{(Left)}~In this example, three presynaptic neurons denoted \textit{b}, \textit{c} and, \textit{d} are fully connected to two post-synaptic neurons \textit{a} and \textit{e}, with different delays of respectively $1$, $5$, and $9~\ms$ for \textit{a} and  $8$, $5$, and $1~\ms$ for \textit{e}. \textit{(Middle)}~If three synchronous pulses are emitted from presynaptic neurons, this will generate post-synaptic potentials that will reach \textit{a} and \textit{e} asynchronously because of the delays, and they may not be sufficient to reach the membrane threshold in either of the post-synaptic neurons, therefore no spike will be emitted.
    %at these different delays, and these may not be sufficient to generate a spike in either neuron.
    \textit{(Right)}~If the pulses are emitted from presynaptic neurons such that, taking into account the delays, these converge at the same time on neuron \textit{a} (here, at $t=10~\ms$) then a spike is emitted (here in red color), while none is emitted on neuron \textit{e}.
